Simulating Ultra High-Dimensional Multivariate Data Using the bigsimr R Package

### 2020-10-16 index.Rmd
migrating ESCO 2020 talk into sections, revising everything

--- 
title: Simulating Ultra High-Dimensional Multivariate Data Using the bigsimr R Package
author:
  - name: Alice Anonymous
    email: alice@example.com
    affiliation: Some Institute of Technology
    footnote: 1
  - name: Bob Security
    email: bob@example.com
    affiliation: Another University
  - name: Cat Memes
    email: cat@example.com
    affiliation: Another University
    footnote: 2
  - name: Derek Zoolander
    email: derek@example.com
    affiliation: Some Institute of Technology
    footnote: 2
address:
  - code: Some Institute of Technology
    address: Department, Street, City, State, Zip
  - code: Another University
    address: Department, Street, City, State, Zip
footnote:
  - code: 1
    text: "Corresponding Author"
  - code: 2
    text: "Equal contribution"
date: "`r format(Sys.Date(), '%A, %B %d, %Y')`"
bibliography:
  - bigsimr.bib
  - packages.bib
site: bookdown::bookdown_site
journal: "Journal of Computational And Applied Mathematics (JCAM)"
#linenumbers: true
#numbersections: true
abstract: |
  In this era of Big Data, it is critical to realistically simulate data to conduct informative Monte Carlo studies. This is problematic when data are inherently multivariate while at the same time are (ultra-) high dimensional. This situation appears frequently in observational data found on online and in high-throughput biomedical experiments (e.g., RNA-sequencing). Due to the difficulty in simulating realistic correlated data points, researchers often resort to simulation designs that posit independence --- greatly diminishing the insight into the empirical operating characteristics of any proposed methodology.  Major challenges lie in the computational complexity involved in simulating these massive random vectors. We propose a fairly general, scalable procedure to simulate high-dimensional multivariate distributions with pre-specified marginal characteristics and dependency characteristics. As a motivating example, we use our methodology to study large-scale statistical inferential procedures applied to cancer-related RNA-sequencing data sets. The proposed algorithm is implemented as the `bigsimr` R package.
---

@@

Along those lines, simulating high-dimensional non-normal data motivates this work --- in pursuit of modeling RNA-sequencing data [@Wang2009b; @Conesa2016b] derived from breast cancer patients. This laboratory procedure results in count data with infinite support, since RNA-sequencing platforms measure gene expression by enumerating the number of reads aligned to genomic regions. Often researchers posit a negative binomial model (refs) as counts are often over (or under) expressed that a Poisson model would suggest. Yet due to inherent biological processes, gene expression data exhibits correlation (coexpression) across genes [@BE07; @Schissler2018]. RNA-sequencing analysis has garnered much interest in the statistical literature (refs). Often researchers simulate independently or from their proposed model in order to conduct Monte Carlo studies (refs). An alternative strategy is to think *generatively* about the data collection process and scientific domain knowledge and design simulations with probabilistic models that reflect those processes. In our motivating example, we'll study the consequences correlation in the simulation study replicating test statistics from an illustrative, classical large-scale hypothesis testing study [see @Efron2004].

@@

To meet the desired properties in the face of the challenging setting, we present a general-purpose, scalable multivariate simulation algorithm. The crux of the method lies in the construction of a Gaussian copula [refs]. The idea that many multivariate and marginally heterogeneous distribution can be constructed in such a manner is well known [refs]. This article's contribution lies in it's application to high-dimensional data through a high-performance implementation (`bigsimr`) and speed-focused algorithm design. The algorithm design relies on useful properties of non-standard correlation measures, namely Kendall's $\tau$ and Spearman's $\rho$. We'll show that quality of simulation does not require the use of the standard Pearson correlation matrix, even for normally distributed data. The purpose of our method is to generate random vectors that match exactly specified any possible Kendall's $\tau$ or Spearman's $\rho$, and approximately Pearson product-moment correlation.

>>>>

--- 
title: Simulating Ultra High-Dimensional Multivariate Data Using the bigsimr R Package
author:
  - name: Alice Anonymous
    email: alice@example.com
    affiliation: Some Institute of Technology
    footnote: 1
  - name: Bob Security
    email: bob@example.com
    affiliation: Another University
  - name: Cat Memes
    email: cat@example.com
    affiliation: Another University
    footnote: 2
  - name: Derek Zoolander
    email: derek@example.com
    affiliation: Some Institute of Technology
    footnote: 2
address:
  - code: Some Institute of Technology
    address: Department, Street, City, State, Zip
  - code: Another University
    address: Department, Street, City, State, Zip
footnote:
  - code: 1
    text: "Corresponding Author"
  - code: 2
    text: "Equal contribution"
date: "`r format(Sys.Date(), '%A, %B %d, %Y')`"
bibliography:
  - bigsimr.bib
  - packages.bib
site: bookdown::bookdown_site
journal: "Journal of Computational And Applied Mathematics (JCAM)"
#linenumbers: true
#numbersections: true
abstract: |
  In this era of Big Data, it is critical to realistically simulate data to conduct informative Monte Carlo studies. This is problematic when data are inherently multivariate while at the same time are (ultra-) high dimensional. This situation appears frequently in observational data found on online and in high-throughput biomedical experiments (e.g., RNA-sequencing). Due to the difficulty in simulating realistic correlated data points, researchers often resort to simulation designs that posit independence --- greatly diminishing the insight into the empirical operating characteristics of any proposed methodology.  Major challenges lie in the computational complexity involved in simulating these massive random vectors. We propose a fairly general, scalable procedure to simulate high-dimensional multivariate distributions with pre-specified marginal characteristics and dependency characteristics. As a motivating example, we use our methodology to study large-scale statistical inferential procedures applied to cancer-related RNA-sequencing data sets. The proposed algorithm is implemented as the bigsimr R package.
---

@@

Along those lines, simulating high-dimensional non-normal data motivates this work --- in pursuit of modeling RNA-sequencing data [@Wang2009b; @Conesa2016b] derived from breast cancer patients. 
This laboratory procedure results in count data with infinite support, since RNA-sequencing platforms measure gene expression by enumerating the number of reads aligned to genomic regions. 
Often researchers posit a negative binomial model (refs) as counts are often over (or under) expressed that a Poisson model would suggest. Yet due to inherent biological processes, gene expression data exhibits correlation (coexpression) across genes [@BE07; @Schissler2018]. 
RNA-sequencing analysis has garnered much interest in the statistical literature (refs). 
Often researchers simulate independently or from their proposed model in order to conduct Monte Carlo studies (refs). 
An alternative strategy is to think *generatively* about the data collection process and scientific domain knowledge and design simulations with probabilistic models that reflect those processes.
Our methodology provides another toolkit to generate data by considering how the data arise in the experimental setting.

@@

To meet the desired properties in the face of the challenging setting, we present a general-purpose, scalable multivariate simulation algorithm. The crux of the method lies in the construction of a Gaussian copula [refs]. The idea that many multivariate and marginally heterogeneous distribution can be constructed in such a manner is well known [refs]. This article's contribution lies in it's application to high-dimensional data through a high-performance implementation (`bigsimr`) and speed-focused algorithm design. The algorithm design relies on useful properties of non-standard correlation measures, namely Kendall's $\tau$ and Spearman's $\rho$. We'll show that quality of simulation does not require the use of the standard Pearson correlation matrix, even for normally distributed data. The purpose of our method is to generate random vectors that match exactly specified any possible Kendall's $\tau$ or Spearman's $\rho$, and approximately Pearson product-moment correlation.More formally, our goal is to simulate $N$ random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with **correlated** components.
The algorithm must scale to a large number of simulation replicates,$N$ for high-dimensional multivariate constructions (dimension $d$ in the tens or hundred thousands.
